---
title: 'BIO00066I Cell Biology '
date: "`r Sys.Date()`"
output:
  html_document: 
    theme: spacelab
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: no
    highlight: tango
  pdf_document:
    toc: yes
  word_document:
    toc: yes
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Step 1: Initial data exploration

--- 

## Setting up
```{r}         

#clear previous data
rm(list=ls())

#The tidyverse!
library(tidyverse)

#we need this to make prerry plots with 'ggarrange'
library(ggpubr)

```

---

## Getting to know the data
Load the data and see what it looks like:
```{r}
load("data/cells.Rda")
ls()
dim(cells)
```


Have a look at the data using **view()**
```{r}
view(cells)
```


Get a summary of all the columns
```{r}
summary(cells)
```


Check how many **tracking ids** and **lineage ids** there are:

```{r}
length(unique(cells$tracking.id))
length(unique(cells$lineage.id))
```

---

## Making plots

It looks like we have these values that could be important to understand.

- volume.um
- radius.um
- width.um
- dry.mass.pgs
- area.um
- length.um

A good way of understanding data is to plot it.

PS: In all these values **um** means microns. These are usually represented as **µm**. 


### Plot a histogram

Histograms can sometimes be useful.Let's see:

```{r}
ggplot(cells, aes(x=volume.um))+
  geom_histogram()
```

### Density plots

A density plot might be be better. So let's try that.

```{r}
ggplot(cells, aes(x=volume.um))+
  geom_density()+
  theme_light()
```

Now examine the **radius** using a **density plot**.


```{r}
ggplot(cells, aes(x=radius.um))+
  geom_density()+
  theme_light()
```
Looks like there might be **two populations** here, perhaps?


### Arranging many plots together
Plot all the values now, **arrange** them and then **save** a plot as a jpeg.


```{r}

#volumne
volume.plot<-ggplot(cells, aes(x=width.um))+
  geom_density()+
  theme_light()

#radius
radius.plot<-ggplot(cells, aes(x=radius.um))+
  geom_density()+
  theme_light()

#width
width.plot<-ggplot(cells, aes(x=width.um))+
  geom_density()+
  theme_light()

#dry.mass.pgs
mass.plot<-ggplot(cells, aes(x=dry.mass.pg))+
  geom_density()+
  theme_light()

#area.um
area.plot<-ggplot(cells, aes(x=area.um))+
  geom_density()+
  theme_light()

#length.um
length.plot<-ggplot(cells, aes(x=length.um))+
  geom_density()+
  theme_light()

#arrange the plot
density.plots<-ggarrange(
  volume.plot,
  radius.plot,
  width.plot, 
  mass.plot, 
  area.plot,
  length.plot,
  ncol = 3, 
  nrow = 2
)

#now save a jpeg
#the default width and height = 7
#so lets make it a bit wider
ggsave(density.plots,
  file="plots/first.density.plots.jpeg",
  width = 7*1.5,
  height = 7
)

```


### Final comments
 
 - What do you see here?
 - What we we wonder if we did **not** do this initial exploration?

### End





